---
title: The conexus package
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The conexus package}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(width=90)
library(terra)
```

## Introduction

*R* package conexus implements methods to support computing the *XC* metric to evaluate the *ex-situ* conservation status of organisms, as described in 

El Graoui, M., M.E. Ghanem, M. Amri, and R.J. Hijmans, 2025. *A distance-based framework for assessing the ex-situ conservation status of plants*. Submitted for publication. 


## Example data 

We first take the conexus and terra packages out of the library, and we read some of the example data. 

```{r data}
library(conexus)
library(terra)

suit <- terra::rast(system.file("ex/suitable.tif", package="conexus"))
land <- terra::readRDS(system.file("ex/land.rds", package="conexus"))
sp <- terra::readRDS(system.file("ex/sp.rds", package="conexus"))
```

`suit` is a "SpatRaster", `land` and `sp` are "SpatVectors". These data types are defined by the "terra" package to handle spatial data. See the tutorials at [rspatial.org](https://rspatial.org) if you are not familiar with "terra". 

`suit` shows where the environment is suitable for our example species *Vigna friesiorum*. This was  predicted with a Species Distribution Model (SDM). We used *R* package "predicts" for that, but we do not show that here. See [rspatial.org/sdm](https://rspatial.org/sdm) for guidance. 

`land` is a SpatVector of polygons delimiting the areas where the species can occur (on land, in this case), and `sp` is a SpatVector of points, showing the locations of know occurrence of the species.

`suit` has values of (`1`) where the environment is suitable for the species according to the SDM, and missing values (`NA`) where it is not suitable. 

To better illustrate the data we first compute a SpatRaster `xrange` that has values of `1` for land areas where the environment is not suitable and `2` for areas where it is suitable.


```{r plotpts, fig.dim=c(6,5)}
rland <- rasterize(land, suit)
xrange <- sum(c(rland, suit), na.rm=TRUE)
plot(xrange, col=c("light gray", "orange"), colNA="azure", mar=c(1,2,1,5),
     levels=c("not suitable", "suitable"), type="classes")
lines(land, lwd=2)
plot(sp, "species", col="blue", plg=list(nudge=c(0,-2), text.font=3), cex=.75, add=TRUE)
``` 

## Range

We use the predictions of suitable areas to estimate the range of the species. We use an "exclusion buffer" to exclude areas that may have a suitable climate, but are too far away from known occurrences to be a credible part of the range. Likewise, we use an "inclusion buffer" to include all areas near known occurrences (after true outliers have been removed, by carefully checking for errors in georeferencing or species identification). 

Here we only show the general area where *V. friesiorum* occurs (mainly in Kenya and Tanzania). But we note that this example comes from a study that studied the distribution of wild *Vigna* species across Africa. The SDM showed environmentally suitable areas in South Africa, Madagascar, and Côte d'Ivoire, all very far away from places where *V. friesiorum* has been observed, and these clearly needed to be removed. 

The `get_range` method takes five arguments. The suitability raster, the point occurrences, land area (or similar) and the inclusion and exclusion buffer sizes (in km). It returns a new SpatRaster. 

 
```{r adjust}
sp_range <- get_range(suit, sp, land, include=50, exclude=250)
sp_range
```

To illustrate this in detail, we first create some new data.


```{r adjdata}
xr <- sum(c(rland, suit, sp_range+1), na.rm=TRUE)
bb <- rbind(terra::buffer(sp, 50000), terra::buffer(sp, 250000))
bb$size <- rep(c("include", "exclude"), each=nrow(sp))
bb <- terra::as.lines(terra::aggregate(bb, "size"))
```

And use these on the map below.

```{r plotadj, fig.dim=c(6,5)}
plot(xr, levels=c("Unsuitable", "Removed", "Added", "Kept"), type="classes", 
      plg=list(title="Range"), mar=c(2,1,1,6), col=c("light gray", map.pal("viridis", 3)))
lines(land, lwd=4, col="light gray")
plot(bb, "size", col=c("darkgreen", "red"), add=TRUE, sort=FALSE, type="classes", lwd=2,
			plg=list(nudge=c(0,-5), title="Buffer", lwd=2))
plot(sp, "species", col="blue", add=TRUE, plg=list(nudge=c(0,-9), text.font=3), cex=.75)
```


## Number of zones

Now that we have a more reasonable estimate of the actual range of the species, we divide it into conservation zones. The number of zones depends on the range size. We use the following (default) function.

```{r fun}
zones <- function(A, omega) pmax(1, round(omega * sqrt(A/pi)))
```

The function takes two arguments. The range size in $km^2$ and a parameter ω (omega). The function can be used like this:

```{r showfun}
range_size <- seq(0, 7500000, 10000)
z <- zones(range_size, 1/40)
head(z, 15)
```

And the plot below shows the effect of changing omega. 

```{r plotfun, fig.dim=c(5,5)}
plot(range_size/1000, z, type="l", xlab="Range size (km^2)", ylab="Number of zones", 
              las=1, col="red", ylim=c(0, 80), lwd=2)
lines(range_size/1000, zones(range_size, 1/20), col="darkgray", lwd=2)
lines(range_size/1000, zones(range_size, 1/80), col="blue", lwd=2)
legend("topleft", legend=c("1/20", "1/40", "1/80"), title="omega",
              lwd=2, col=c("darkgray", "red", "blue"))
```

We use `get_samplesize` with the default function above, to compute the sample-size (number of zones) for the species. 

```{r ssize}
k <- get_samplesize(sp_range, omega=1/40)

str(k)
```

Output `k` is a list of with three elements: "range" is a SpatVector of polygons for the range, "area" is the range size in $km^2$, and "n" is the suggested number of zones.  


## Make zones

Now that we have the number of zones we want, we need to create them. We can do this with the `make_zones` function. It requires a SpatRaster with the x and y coordinates of the area, the species range and the number of zones to create. Under the hood, it uses *k*-means clustering. Argument "spread" assures that the initial cluster centers are well spread out. This is important because they are assigned at random, and they can sometimes get stuck in sub-optimal places. 

```{r zones}
xy <- c(terra::init(sp_range, "x"), terra::init(sp_range, "y"))
set.seed(321)
zones <- make_zones(xy, k$range, k$n, spread=TRUE)
```

Above we used `set.seed` because *k*-means clustering is stochastic. That is, each time you run it, it will give somewhat different results, unless you initialize the random number generator to a fixed value before you run it.

The output from `make_zones` is a SpatVector of polygons.

```{r plotzones, fig.dim=c(5,5)}
plot(sp_range, legend=F)
lines(zones, col="red", lwd=2)
text(zones, halo=TRUE)
```



## Geo-environmental distance 

The *XC* metric depends on the distance between zones. This distance has a geographic element: the further apart zones are, the more isolated populations in these area are from each other. But it also has an environmental aspect: populations in very different environments may have evolved to be more genetically different than expected by geographic distance alone. 

To combine geographic and environmental distance into a single distance metric, we express environmental distance as a function of geographic distance. We computed this relationship using all *Vigna* occurrences in Africa. Below we show the data and modeled relationship used. We used annual precipitation and annual mean temperature as environmental variables. 

```{r envd, fig.dim=c(7,5)}
mprc <- readRDS(system.file("ex/m_prc.rds", package="conexus"))
mtmp <- readRDS(system.file("ex/m_tmp.rds", package="conexus"))

par(mfrow=c(1,2))
plot(mprc, cex=.5, xlab="Δ precipiation (mm)", ylab="geographic distance", col="gray")
prc=seq(0,2500,100)
lines(prc, predict(mprc, data.frame(prc=prc)), col="blue", lwd=2)
plot(mtmp, cex=.5, xlab="Δ temperature (°C)", ylab="geographic distance", col="gray")
tmp=seq(0,15,1)
lines(tmp, predict(mtmp, data.frame(tmp=tmp)), col="red", lwd=2)
```

We can write a function that computes the expected minimum geographic distance between two points given the environmental distance. We cap this at the maximum distance observed -- this is not strictly necessary, as this situation will not occur for any particular species). We average the expected distancew for the two environmental variables (temperature and precipitation).


```{r}
envdist <- function(x) {
	x$tmp[x$tmp > 13] <- 13 
	x$prc[x$prc > 2000] <- 2000
	p_tmp <- predict(mtmp, x)
	p_pr <- predict(mprc, x)
	rowMeans(cbind(p_tmp, p_pr))
}
```


This function is used in the next step.


## Conservation score

We can now compute the *XC* score with the `XC` function. We need to provide the environmental data (temperature and precipitation) to match the geo-environmental distance function that we developed above.

```{r envdata}
env <- terra::rast(system.file("ex/env.tif", package="conexus"))
```

`XC` has seven parameters. It needs the conservation zones (polygons) and the locations where seed samples were collected. To use the geo-environmental distance, the environmental SpatRaster the function that transforms environmental distance to geographic distance need to be provided. Otherwise, only the geographic distance is used. In addition there are two parameter arguments "minssize", that sets the minimal sample size required for a perfect XC score of 1; and "maxdist" argument that allows capping the distance between zones (this can be useful for highly disjoint ranges).

```{r XC}
seed <- sp[sp$sample_type=="seed"]
x <- XC(zones, seed, env=env, envfun=envdist, minssize=10, maxdist=1500)
x$XC 
```

The *XC* conservation score is `r round(x$XC, 3)`. 

The information on the network is also returned and can be used to make a map.

```{r xd, fig.dim=c(6,5)}
head(round(x$dist, 2))
v <- dst2svect(x$dist)
xctr <- centroids(zones, inside=T)
zones$seed <- "no seed"
zones$seed[extract(zones, seed)[,2]] <- "seed"
plot(zones, "seed", col=c("azure","lightyellow"), border="lightgray",
     mar=c(2,1,1,8), plg=list(title="Zones"))
lines(v, col="red", lwd=5*v$weight/1000)
points(xctr, col="gray", cex=3)
points(xctr[10], col="lightgreen", cex=3)
text(xctr, halo=TRUE, col="blue", font=2)
legend(40.8,3,c("network\n(edge weight)"), lwd=6, col="red", xpd=TRUE, cex=.8, box.lty=0)

```

To further illustrate the network approach we can compare the network edge weights with no seed sample, with the edge weights with seed samples from zones 2, 5, 7, and 10. 

```{r dist, fig.dim=c(6,5)}
# no sample
x <- XC(zones, seed[0,], env=env, envfun=envdist)
x$XC
# four seed samples, from four zones
seed <- sp[c(1,8,33, 34),]
y <- XC(zones, seed, env=env, envfun=envdist)
y$XC
zones$seed <- "no seed"
zones$seed[extract(zones, seed)[,2]] <- "seed"

# make SpatVectors
vx <- dst2svect(x$dist)
vy <- dst2svect(y$dist)
# identify the zones with seed samples
# get centroids
xctr <- centroids(zones, inside=TRUE)
sctr <- centroids(zones[zones$seed=="seed"], inside=TRUE)

plot(zones, "seed", col=c("azure", "lightyellow"), border="lightgray",
     mar=c(2,1,1,8), plg=list(title="Zones"))
lines(vx, col="red", lwd=vx$weight/100)
i <- !is.na(vy$envdst) & vy$w!=1
lines(vy[i], col="blue", lwd=vy$weight[i]/100)
lines(vy[is.na(vy$envdst)], col="blue", lty=3, lwd=2)
points(xctr, col=gray(0.99), cex=3)
points(sctr, col="lightgreen", cex=3)
text(xctr, halo=TRUE, font=2)
legend(40.8,3,c("original", "changed", "new"), lwd=c(9,6,3), col=c("red", "blue", "blue"), 
       lty=c(1,1,3), xpd=TRUE, cex=.8, box.lty=0, title="Network connections\nand weights")
``` 

The original network is shows as red lines, their width is proportional to the edge weights (the geo-environmental distance). The edges that have a different weight when considering the four zones with seed samples are shown in blue. If one of two adjacent zones was sampled the edge weight is half the original weight (e.g. all connections between zone 10 and its neighbors). If two  adjacent zones were both sampled, their edge-weight becomes zero (there is one case, the edge between zones 2 and 5). There are also new new "shortcut" edges with zero weight that connect sampled zones that were previously unconnected (for example, zone 10 is now connected to zones 2, 5, and 7). 

## Low sample size penalty 

The *XC* score for the sample described above is `r round(y$XC, 3)`. This is in part because the seed sample size is only four, so there is a penalty for the low sample size. Without that penalty

```{r ssizepen}
z <- XC(zones, seed, env=env, envfun=envdist, minssize=1)
z$XC
```
the *XC* score is `r z$XC`.

The sample size penalty is currently implemented like this 

```{r penalty, fig.dim=c(5,5)}
conexus:::small_ssize_penalty

ssize <- 0:15
penalty <- conexus:::small_ssize_penalty(ssize, 10)
plot(penalty~ssize, type="b", xlab="Sample size", ylab="1-penalty", las=1, col="red", lwd=2)
```

The original *XC* is multiplied with this penalty (1 is no penalty).


## Non-georeferenced seed samples

To adjust for non-georeferenced seed samples a function of the expected value of XC given the number of seed samples can be created. In our paper we use a LOESS function that was somewhat similar to the one shown below. 

```{r, fig.dim=c(5,5)}
expected_XC <- function(x, a=0.79, b=0.28, c=9.5) {
  a / (1 + exp(-b * (x - c)))
}

x <- 1:30
plot(x, expected_XC(x), xlab="n seed samples", ylab="expected XC", type="l", las=1, lwd=2, col="blue")
```

We propose to use the maximum value of the XC computed above, and the expected XC. For example, if you had XC of 0.2 and 7 non-georeferenced seed samples the adjusted XC would become `r round(expected_XC(7), 2)`.

```{r}
expected_XC(7)
```

